---
title: "England & Wales Correlation Heatmap"
author: "Zain Asaf"
date: "2025-02-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}

library (tidyverse)
library(ggplot2)
library(here)

```

```{r}

census_df <- read_csv(here("data/e&w_census_data.csv")) 
segmentation_df <- read_csv(here("data/england_and_wales_oa_segmentation.csv"))

```

```{r}

census_df <- census_df %>% rename(oa21cd = OA21CD)

# Merge datasets on 'oa21cd'
merged_df_england_wales <- left_join(census_df, segmentation_df, by = "oa21cd")

numeric_cols <- merged_df_england_wales %>% select(where(is.numeric))

# Compute Pearson correlation matrix
cor_matrix <- cor(numeric_cols, use = "pairwise.complete.obs", method = "pearson")

# Extract correlations between demographics and PCA dimensions
pca_correlation <- cor_matrix %>%
  as_tibble(rownames = "variable") %>%
  select(variable, low_diversity, security) %>%
  filter(!variable %in% c("low_diversity", "security"))

# Save the correlation matrix as CSV
write_csv(pca_correlation, here("data/demographic_pca_correlations_e&w.csv"))
write_csv(merged_df_england_wales,here("data/merged_df_england_wales.csv"))


# Print the first few rows
print(head(pca_correlation))
```

```{r}

pca_correlation_long <- pca_correlation %>%
  pivot_longer(cols = c(low_diversity, security), names_to = "dimension", values_to = "correlation")

pca_correlation_long <- pca_correlation_long %>%
  mutate(variable = reorder(variable, abs(correlation)))

```

```{r}

pca_correlation <- pca_correlation %>%
  rename(RowIndex = `...1`)  

```



```{r}
variable_categories <- tibble(
  variable = pca_correlation$variable,
  category = case_when(
    str_detect(variable, "occupation|Manager|Process") ~ "Occupation",  
    str_detect(variable, "qualifications|Apprenticeship") ~ "Qualifications",
    str_detect(variable, "Owned|Shared|rented|Lives") ~ "Housing Tenure",
    str_detect(variable, "Household_") ~ "Household Deprivation Status",
    str_detect(variable, "Single_family_household|One_person_household|Multi_person_household|Other_household_types") ~ "Household Type",  
    str_detect(variable, "Male|Female") ~ "Sex",
    str_detect(variable, "national|identity|identities") ~ "National Identity",
    str_detect(variable, "Asian|Black|Mixed|White|Other_ethnic|Any_other_ethnic") ~ "Ethnicity",
    str_detect(variable, "Economically") ~ "Economic Status",
    str_detect(variable, "Europe|Africa|Middle_East|The_Americas|British_Overseas|Antarctica") ~ "Place of Birth",
    str_detect(variable, "Married|Divorced|Separated|Widowed|Never_married") ~ "Marital Status",
    str_detect(variable, "deprivation") ~ "Household Deprivation Status",
    str_detect(variable, "Welsh") ~ "Welsh Language",
    str_detect(variable, "Not_in") ~ "Employment Status",
    str_detect(variable, "health") ~ "Health Status", 
    str_detect(variable, "Christian|Buddhist|Hindu|Jewish|Muslim|Sikh|Other_religion|Not_answered|No_religion") ~ "Religion",
    str_detect(variable, "All_houshold_members|Ethnic_groups_differ|_same_ethnic_group") ~ "Household Ethnic Composition",
    str_detect(variable, "segment") ~ "Segment",
    str_detect(variable, "Aged") ~ "Age",  # Moved this below "Household Type"
    TRUE ~ "Other"
  )
)


```

```{r}

# Merge category labels with correlation data
pca_correlation_long <- pca_correlation %>%
  pivot_longer(cols = c(low_diversity, security), names_to = "dimension", values_to = "correlation") %>%
  left_join(variable_categories, by = "variable")

write_csv(pca_correlation_long, ("data/demographic_pca_correlations_long_e&w.csv"))

```



```{r}

pca_correlation_long_filtered_1 <- pca_correlation_long %>%
  filter(category %in% c("Age", "Ethnicity", "Health Status", "Sex", "Qualifications", "Housing Tenure", "Occupation", "Economic Satus")) 

plot_filtered_1 <- ggplot(pca_correlation_long_filtered_1, aes(x = dimension, y = reorder(variable, correlation), fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), color = "black", size = 3) +  # Add numeric values
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Correlation Heatmaps: Demographics vs PCA Dimensions England & Wales",
       x = "PCA Dimension", y = "Demographic Variable", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~category, scales = "free_y")  # Separate heatmaps for each category


print(plot_filtered_1)

```

```{r}
pca_correlation_long_filtered_2 <- pca_correlation_long %>%
  filter(category %in% c("Place of Birth", "National Idenitity", "Marital Status", "Welsh Language", "Employment Status", "Health Status", "Religion", "Segment", "Household Type")) 


plot_filtered_2 <- ggplot(pca_correlation_long_filtered_2, aes(x = dimension, y = reorder(variable, correlation), fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), color = "black", size = 3) +  # Add numeric values
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Correlation Heatmaps: Demographics vs PCA Dimensions England & Wales",
       x = "PCA Dimension", y = "Demographic Variable", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~category, scales = "free_y")  # Separate heatmaps for each category


print(plot_filtered_2)


