---
title: "Untitled"
author: "Zain Asaf"
date: "2025-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(tidyverse)
library(ggplot2)
```

```{r}
census_df <- read_csv("data/OA21CD_census_2021.csv") 
segmentation_df <- read_csv("data/scotland_oa_segmentation.csv")
```

```{r}
census_df <- census_df %>% rename(oa21cd = OA21CD)

# Merge datasets on 'oa21cd'
merged_df_scotland <- left_join(census_df, segmentation_df, by = "oa21cd")

numeric_cols <- merged_df_scotland %>% select(where(is.numeric))

# Compute Pearson correlation matrix
cor_matrix <- cor(numeric_cols, use = "pairwise.complete.obs", method = "pearson")

# Extract correlations between demographics and PCA dimensions
pca_correlation <- cor_matrix %>%
  as_tibble(rownames = "variable") %>%
  select(variable, low_diversity, security) %>%
  filter(!variable %in% c("low_diversity", "security"))


write_csv(pca_correlation, ("data/demographic_pca_correlations.csv"))
write_csv(merged_df_scotland, ("data/merged_df_scotland.csv"))

# Print the first few rows
print(head(pca_correlation))

```

```{r}



pca_correlation_long <- pca_correlation %>%
  pivot_longer(cols = c(low_diversity, security), names_to = "dimension", values_to = "correlation")


pca_correlation_long <- pca_correlation_long %>%
  mutate(variable = reorder(variable, abs(correlation)))

# Plot bar chart
ggplot(pca_correlation_long, aes(x = variable, y = correlation, fill = dimension)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +  # Flip for readability
  labs(title = "Correlation of Demographics with PCA Dimensions",
       x = "Demographic Variable", y = "Pearson Correlation") +
  theme_minimal()


```

```{r}
## sorted by correlation value

pca_correlation_long <- pca_correlation %>%
  pivot_longer(cols = c(low_diversity, security), 
               names_to = "dimension", values_to = "correlation")


ggplot(pca_correlation_long, aes(x = dimension, y = reorder(variable, correlation), fill = correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Correlation Heatmap: Demographics vs PCA Dimensions",
       x = "PCA Dimension", y = "Demographic Variable", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```

```{r}

ggplot(pca_correlation_long, aes(x = variable, y = correlation, color = dimension)) +
  geom_point(size = 4) +
  geom_segment(aes(x = variable, xend = variable, y = 0, yend = correlation)) +
  coord_flip() +
  labs(title = "Lollipop Chart: Demographics vs PCA Correlation",
       x = "Demographic Variable", y = "Pearson Correlation") +
  theme_minimal()

```

```{r}
## best way to do it is with the nine different categoreis 
## 1) age
##  2) qualifcation 
## 3) sex 
##4)  national identity 
## 5) ethnicity 
## 6) tenusere 
## 7) marital status 
## 8) economic activity 
## 9) occupation 
## 10 ) accomodation 
## 11) distance travelled to work 
## 12) general health
## 13) household-cars
## 14) houshold cars
## 15) religion 
```

```{r}

# Define categories with more precise ordering
variable_categories <- tibble(
  variable = pca_correlation$variable,
  category = case_when(
    str_detect(variable, "tenure") ~ "Housing Tenure",  
    str_detect(variable, "occupation") ~ "Occupation",  
    str_detect(variable, "^age_") ~ "Age",  
    str_detect(variable, "qualification") ~ "Qualification",
    str_detect(variable, "distance") ~ "Distance Travelled to Work",
    str_detect(variable, "sex") ~ "Sex",
    str_detect(variable, "national") ~ "National Identity",
    str_detect(variable, "ethnic") ~ "Ethnicity",
    str_detect(variable, "marital") ~ "Marital Status",
    str_detect(variable, "accommodation") ~ "Accommodation Type",
    str_detect(variable, "general") ~ "Health Status", 
    str_detect(variable, "household") ~ "Number of Cars in Household",
    str_detect(variable, "travel") ~ "Mode of Travel to Work",
    str_detect(variable, "residence") ~ "Length of Time in UK",
    str_detect(variable, "disability") ~ "Level of Disability", 
    str_detect(variable, "unpaid") ~ "Care Status",           
    str_detect(variable, "religion") ~ "Religion",
    str_detect(variable, "economic") ~ "Economic Status",
    str_detect(variable, "segment") ~ "Segment",
    TRUE ~ "Other"
  )
)

```

```{r}
# Merge category labels with correlation data
pca_correlation_long <- pca_correlation %>%
  pivot_longer(cols = c(low_diversity, security), names_to = "dimension", values_to = "correlation") %>%
  left_join(variable_categories, by = "variable")



write_csv(pca_correlation_long, ("data/demographic_pca_correlations_long_scotland.csv"))

```

```{r}

pca_correlation_long_filtered_1 <- pca_correlation_long %>%
  filter(category %in% c("Age", "Ethnicity", "Health Status", "Sex", "Qualification", "Housing Tenure", "Occupation", "Economic Satus")) 

plot_filtered_1 <- ggplot(pca_correlation_long_filtered_1, aes(x = dimension, y = reorder(variable, correlation), fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), color = "black", size = 3) +  # Add numeric values
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Correlation Heatmaps: Demographics vs PCA Dimensions: Scotland",
       x = "PCA Dimension", y = "Demographic Variable", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~category, scales = "free_y")  # Separate heatmaps for each category


print(plot_filtered_1)

```

```{r}

pca_correlation_long_filtered_1 <- pca_correlation_long %>%
  filter(category %in% c("Age", "Ethnicity", "Health Status", "Sex", "Qualification", "Housing Tenure", "Occupation", "Economic Satus")) 

plot_filtered_1 <- ggplot(pca_correlation_long_filtered_1, aes(x = dimension, y = reorder(variable, correlation), fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), color = "black", size = 3) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Correlation Heatmaps: Demographics vs PCA Dimensions: Scotland",
       x = "PCA Dimension", y = "Demographic Variable", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~category, scales = "free_y")  # Separate heatmaps for each category


print(plot_filtered_1)

```


```{r}
pca_correlation_long_filtered_2 <- pca_correlation_long %>%
  filter(category %in% c("National Identity", "Marital Satus", "Accommodation Type", "Mode of Travel to Work", "Level of Disability", "Care Status", "Segment")) 

plot_filtered_2 <- ggplot(pca_correlation_long_filtered_2, aes(x = dimension, y = reorder(variable, correlation), fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), color = "black", size = 3) +  
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, limits = c(-1, 1))  +
  labs(title = "Correlation Heatmaps: Demographics vs PCA Dimensions: Scotland",
       x = "PCA Dimension", y = "Demographic Variable", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~category, scales = "free_y")  # Separate heatmaps for each category


print(plot_filtered_2)





```

# Create heatmaps for each category
ggplot(pca_correlation_long, aes(x = dimension, y = reorder(variable, correlation), fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), color = "black", size = 3) +  # Add numeric values
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Correlation Heatmaps: Demographics vs PCA Dimensions",
       x = "PCA Dimension", y = "Demographic Variable", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~category, scales = "free_y")  # Separate heatmaps for each category

```

