---
title: "Segment_Demo Comparison"
author: "Zain Asaf"
date: "2025-02-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r} 
library(tidyverse)
library(ggplot2)
library(readr)
```

```{r}
scotland <- read_csv("data/merged_df_scotland.csv")

england_wales <- read_csv("data/merged_df_england_wales.csv")

```

```{r}

library(tidyverse)

# Preprocess the England/Wales data
england_wales <- england_wales %>%
  rename(
    # Rename columns if necessary (based on your data structure)
    # For example:
    # sex_female = Female,
    # sex_male = Male
  ) %>%
  mutate(
    age_0_15 = Aged_4_years_and_under + Aged_5_to_9_years + Aged_10_to_14_years,
    age_16_24 = Aged_15_to_19_years + Aged_20_to_24_years,
    age_25_34 = Aged_25_to_29_years + Aged_30_to_34_years,
    age_35_49 = Aged_35_to_39_years + Aged_40_to_44_years + Aged_45_to_49_years,
    age_50_64 = Aged_50_to_54_years + Aged_55_to_59_years + Aged_60_to_64_years,
    age_65_and_over = Aged_65_to_69_years + Aged_70_to_74_years + Aged_75_to_79_years + Aged_80_to_84_years + Aged_85_years_and_over
  )

# Print the first few rows of the cleaned data to verify
head(england_wales)

```

```{r}


segment_num <- 6

# Filter for the specified segment
england_wales_segment <- england_wales %>%
  filter(segment == segment)

# Pivot and summarize
england_wales_summary <- england_wales_segment %>%
  pivot_longer(cols = -c(oa21cd, segment, segment_name),  # Convert all demographic vars into long format
               names_to = "category", 
               values_to = "value") %>%
  group_by(category) %>%
  summarize(mean_value = mean(value, na.rm = TRUE)) %>%
  ungroup()

# View output
print(england_wales_summary)

```












```{r}

# Create the variable_categories tibble
variable_categories <- tibble(
  variable = colnames(england_wales),
  category = case_when(
    str_detect(variable, "occupation|Manager|Process") ~ "Occupation",  
    str_detect(variable, "qualifications|Apprenticeship") ~ "Qualifications",
    str_detect(variable, "Owned|Shared|rented|Lives") ~ "Housing Tenure",
    str_detect(variable, "Household_") ~ "Household Deprivation Status",
    str_detect(variable, "Single_family_household|One_person_household|Multi_person_household|Other_household_types") ~ "Household Type",  
    str_detect(variable, "Male|Female") ~ "Sex",
    str_detect(variable, "national|identity|identities") ~ "National Identity",
    str_detect(variable, "Asian|Black|Mixed|White|Other_ethnic|Any_other_ethnic") ~ "Ethnicity",
    str_detect(variable, "Economically") ~ "Economic Status",
    str_detect(variable, "Europe|Africa|Middle_East|The_Americas|British_Overseas|Antarctica") ~ "Place of Birth",
    str_detect(variable, "Married|Divorced|Separated|Widowed|Never_married") ~ "Marital Status",
    str_detect(variable, "deprivation") ~ "Household Deprivation Status",
    str_detect(variable, "Welsh") ~ "Welsh Language",
    str_detect(variable, "Not_in") ~ "Employment Status",
    str_detect(variable, "health") ~ "Health Status", 
    str_detect(variable, "Christian|Buddhist|Hindu|Jewish|Muslim|Sikh|Other_religion|Not_answered|No_religion") ~ "Religion",
    str_detect(variable, "All_houshold_members|Ethnic_groups_differ|_same_ethnic_group") ~ "Household Ethnic Composition",
    str_detect(variable, "segment") ~ "Segment",
    str_detect(variable, "age") ~ "Age",  # Moved this below "Household Type"
    TRUE ~ "Other"
  )
)

```


```{r}


england_wales_long <- england_wales %>%
  pivot_longer(
    cols = -c(oa21cd, segment, segment_name),  # Exclude both numeric 'segment' and character 'segment_name'
    names_to = "variable",
    values_to = "value"
  )


```




# Convert Scotland data to long format
scotland_long <- scotland_clean %>%
  pivot_longer(
    cols = -c(segment),  # Exclude segment from pivoting
    names_to = "variable",
    values_to = "value"
  )

# Join the long data with variable_categories to add the category column
england_wales_long <- england_wales_long %>%
  left_join(variable_categories, by = "variable")

scotland_long <- scotland_long %>%
  left_join(variable_categories, by = "variable")

# Define the segment number
segment_num <- 6

# Filter the data for the specified segment
england_wales_segment <- england_wales_long %>%
  filter(segment == segment_num)

scotland_segment <- scotland_long %>%
  filter(segment == segment_num)

# Summarize the demographic variables for the specified segment in England/Wales
england_wales_summary <- england_wales_segment %>%
  group_by(category, variable) %>%
  summarize(
    mean_value = mean(value, na.rm = TRUE)
  ) %>%
  ungroup()

# Summarize the demographic variables for the specified segment in Scotland
scotland_summary <- scotland_segment %>%
  group_by(category, variable) %>%
  summarize(
    mean_value = mean(value, na.rm = TRUE)
  ) %>%
  ungroup()

# Plot for England/Wales
ggplot(england_wales_summary, aes(x = variable, y = mean_value, fill = category)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Demographic Summary for Segment 6 in England/Wales", x = "Demographic Variable", y = "Mean Value") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Plot for Scotland
ggplot(scotland_summary, aes(x = variable, y = mean_value, fill = category)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Demographic Summary for Segment 6 in Scotland", x = "Demographic Variable", y = "Mean Value") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))









```

```{r}



# Preprocess the England/Wales data
england_wales_clean <- england_wales %>%
  rename(
    oa21cd = oa21cd,
   # segment = `...1`,
    sex_female = Female,
    sex_male = Male
  ) %>%
  mutate(
    age_0_15 = Aged_4_years_and_under + Aged_5_to_9_years + Aged_10_to_14_years,
    age_16_24 = Aged_15_to_19_years + Aged_20_to_24_years,
    age_25_34 = Aged_25_to_29_years + Aged_30_to_34_years,
    age_35_49 = Aged_35_to_39_years + Aged_40_to_44_years + Aged_45_to_49_years,
    age_50_64 = Aged_50_to_54_years + Aged_55_to_59_years + Aged_60_to_64_years,
    age_65_and_over = Aged_65_to_69_years + Aged_70_to_74_years + Aged_75_to_79_years + Aged_80_to_84_years + Aged_85_years_and_over
  )




```

```{r}

# Filter data for the specified segment (e.g., segment 6)
segment_num <- 6
england_wales_segment <- england_wales_clean %>% filter(segment == segment_num)
scotland_segment <- scotland_clean %>% filter(segment == segment_num)

# Summarize the demographic variables for the specified segment in England/Wales
england_wales_summary <- england_wales_segment %>%
  summarize(
    sex_female = mean(sex_female, na.rm = TRUE),
    sex_male = mean(sex_male, na.rm = TRUE),
    age_0_15 = mean(age_0_15, na.rm = TRUE),
    age_16_24 = mean(age_16_24, na.rm = TRUE),
    age_25_34 = mean(age_25_34, na.rm = TRUE),
    age_35_49 = mean(age_35_49, na.rm = TRUE),
    age_50_64 = mean(age_50_64, na.rm = TRUE),
    age_65_and_over = mean(age_65_and_over, na.rm = TRUE)
  )

# Summarize the demographic variables for the specified segment in Scotland
scotland_summary <- scotland_segment %>%
  summarize(
    sex_female = mean(sex_female, na.rm = TRUE),
    sex_male = mean(sex_male, na.rm = TRUE),
    age_0_15 = mean(age_0_15, na.rm = TRUE),
    age_16_24 = mean(age_16_24, na.rm = TRUE),
    age_25_34 = mean(age_25_34, na.rm = TRUE),
    age_35_49 = mean(age_35_49, na.rm = TRUE),
    age_50_64 = mean(age_50_64, na.rm = TRUE),
    age_65_and_over = mean(age_65_and_over, na.rm = TRUE)
  )

# Combine the summaries for comparison
comparison_result <- england_wales_summary %>%
  bind_rows(scotland_summary) %>%
  mutate(region = c("England/Wales", "Scotland"))

# Print the comparison result
print(comparison_result)