---
title: "Segment Demo Comprison"
author: "Zain Asaf"
date: "2025-02-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(tidyverse)
library(purrr)

```

```{r}
 england_wales <- readRDS("data/england_wales_rename.rds")

  scotland <- read.csv("data/merged_df_scotland.csv")
```

```{r}

# Group by summarise mean

data_joined <- full_join(
 
  england_wales %>%
    group_by(segment) %>%
    summarise(across(where(is.numeric), mean, na.rm = TRUE)),

  scotland %>%
    group_by(segment) %>%
    summarise(across(where(is.numeric), mean, na.rm = TRUE)),
 
  by="segment")

write.csv(data_joined, "data/data_joined.csv")


```

```{r}

# Function to subtract .y variables from .x variables
subtract_xy <- function(df) {
  # Get all column names with .x and .y suffixes
  x_columns <- grep("\\.x$", colnames(df), value = TRUE)
  y_columns <- grep("\\.y$", colnames(df), value = TRUE)
  
  # Remove the suffix to get base column names
  base_names <- sub("\\.x$", "", x_columns)
  
  # Create a new data frame with the differences
  difference_df <- df %>%
    select(segment, all_of(x_columns), all_of(y_columns)) %>%
    mutate(across(all_of(x_columns), 
                  ~ . - df[[sub("\\.x$", ".y", cur_column())]], 
                  .names = "difference_{col}")) %>%
    select(segment, starts_with("difference_"))
  
  # Rename the columns to remove the "difference_" prefix
  colnames(difference_df) <- c("segment", base_names)
  
  return(difference_df)
}

# Apply the function to your data frame
data_joined_diff <- subtract_xy(data_joined)


write.csv(data_joined_diff, "data/data_joined_diff.csv")

```

```{r}

numeric_cols <- data_joined_diff %>% select(-c(segment)) %>% names()

# Function to get top 3 absolute values for a variable
get_top_segments <- function(var) {
  data_joined_diff %>%
    select(segment, all_of(var)) %>%
    arrange(desc(abs(!!sym(var)))) %>%
    slice_head(n = 3) %>%
    mutate(variable = var)  
}

# Apply function to each numeric column
top_segments_list <- map(numeric_cols, get_top_segments)

# Combine results into a single dataframe
top_segments_df <- bind_rows(top_segments_list) %>%
  select(variable, segment, everything())  

# View results
print(top_segments_df)

```




```{r}



top_segments_demographic <- data_joined_diff %>% 
  pivot_longer(cols = -segment, names_to = "variable", values_to = "value") %>% 
  mutate(demographic_group = case_when(
    str_detect(variable, "sex") ~ "Sex",
    str_detect(variable, "age") ~ "Age",
    str_detect(variable, "qualification") ~ "Qualification",
    str_detect(variable, "tenure") ~ "Housing Tenure",
    str_detect(variable, "ethnicity") ~ "Ethnicity",
    str_detect(variable, "occupation") ~ "Occupation",
    str_detect(variable, "economic_activity") ~ "Economic Activity",
    str_detect(variable, "qualifications") ~ "Qualifications",
    str_detect(variable, "economic") ~ "Economic Activity",
    str_detect(variable, "religion") ~ "Religion",
    TRUE ~ "Other"
  ))


head(top_segments_demographic)


top_segments_demographic <- top_segments_demographic %>% 
  mutate(abs_value = abs(value)) %>%  
  group_by(demographic_group) %>% 
  slice_max(order_by = abs_value, n = 3) %>%  
  ungroup()  

# View the result
print(top_segments_demographic)



write.csv(top_segments_demographic, "data/top_segments_demographic.csv")

