---
title: "Data Cleaning Segment Comparison"
author: "Zain Asaf"
date: "2025-02-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r} 
library(tidyverse)
library(stringr)
```

```{r} 

scotland <- read_csv("data/merged_df_scotland.csv")

england_wales <- read_csv("data/merged_df_england_wales.csv")

```

```{r}
## remove scientific notation and 

options(scipen = 999)  



```

```{r}
# Recode age variables in England & Wales to match Scotland's categories
england_wales <- england_wales %>%
  mutate(
    age__0_15 = Aged_4_years_and_under + Aged_5_to_9_years + Aged_10_to_14_years,
    age__16_24 = Aged_15_to_19_years + Aged_20_to_24_years,
    age__25_34 = Aged_25_to_29_years + Aged_30_to_34_years,
    age__35_49 = Aged_35_to_39_years + Aged_40_to_44_years + Aged_45_to_49_years,
    age__50_64 = Aged_50_to_54_years + Aged_55_to_59_years + Aged_60_to_64_years,
    age__65_and_over = Aged_65_to_69_years + Aged_70_to_74_years + Aged_75_to_79_years + 
                       Aged_80_to_84_years + Aged_85_years_and_over
  )

```

```{r}

## define function for england ## 

calculate_normalized_proportions_england <- function(data) {
  data_long <- data %>%
    pivot_longer(cols = -segment, names_to = "variable", values_to = "count") %>%  # Reshape first
    mutate(category = case_when(
      str_detect(variable, "occupation|Manager|Process") ~ "Occupation",  
      str_detect(variable, "qualifications|Apprenticeship") ~ "Qualifications",
      str_detect(variable, "Owned|Shared|rented|Lives") ~ "Housing Tenure",
      str_detect(variable, "Household_") ~ "Household Deprivation Status",
      str_detect(variable, "Single_family_household|One_person_household|Multi_person_household|Other_household_types") ~ "Household Type",  
      str_detect(variable, "Male|Female") ~ "Sex",
      str_detect(variable, "national|identity|identities") ~ "National Identity",
      str_detect(variable, "Asian|Black|Mixed|White|Other_ethnic|Any_other_ethnic") ~ "Ethnicity",
      str_detect(variable, "Economically") ~ "Economic Status",
      str_detect(variable, "Europe|Africa|Middle_East|The_Americas|British_Overseas|Antarctica") ~ "Place of Birth",
      str_detect(variable, "Married|Divorced|Separated|Widowed|Never_married") ~ "Marital Status",
      str_detect(variable, "deprivation") ~ "Household Deprivation Status",
      str_detect(variable, "Welsh") ~ "Welsh Language",
      str_detect(variable, "Not_in") ~ "Employment Status",
      str_detect(variable, "health") ~ "Health Status", 
      str_detect(variable, "Christian|Buddhist|Hindu|Jewish|Muslim|Sikh|Other_religion|Not_answered|No_religion") ~ "Religion",
      str_detect(variable, "All_houshold_members|Ethnic_groups_differ|_same_ethnic_group") ~ "Household Ethnic Composition",
      str_detect(variable, "segment") ~ "Segment",
      str_detect(variable, "Aged") ~ "Age",
      TRUE ~ "Other"
    )) %>%
    group_by(segment, category) %>% 
    summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop") %>%  segment-category
    group_by(segment, category) %>% 
    mutate(
      proportion = (total_count / sum(total_count)) * 100,  
      proportion = round(proportion, 2),
      proportion_formatted = paste0(proportion, "%")  
    ) %>%
    select(segment, category, variable, proportion, proportion_formatted)  

  return(data_long)
}


```

```{r}
england_summary <- calculate_normalized_proportions_england(england_wales) %>% mutate(region = "England & Wales")


```

```{r}

## define function for Scotland 

calculate_normalized_proportions_scotland <- function(data) {
  data_long <- data %>%
    group_by(segment) %>%
    summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>%
    pivot_longer(cols = -segment, names_to = "variable", values_to = "count") %>%
    mutate(category = case_when(
      str_detect(variable, "tenure") ~ "Housing Tenure",  
      str_detect(variable, "occupation") ~ "Occupation",  
      str_detect(variable, "age__") ~ "Age",  
      str_detect(variable, "qualifications") ~ "Qualification",
      str_detect(variable, "distance_to_work") ~ "Distance Travelled to Work",
      str_detect(variable, "sex__") ~ "Sex",
      str_detect(variable, "national_identity") ~ "National Identity",
      str_detect(variable, "ethnicity__") ~ "Ethnicity",
      str_detect(variable, "marital_status") ~ "Marital Status",
      str_detect(variable, "accommodation_type") ~ "Accommodation Type",
      str_detect(variable, "general_health") ~ "Health Status", 
      str_detect(variable, "cars_in_household") ~ "Number of Cars in Household",
      str_detect(variable, "travel_mode") ~ "Mode of Travel to Work",
      str_detect(variable, "residence_length") ~ "Length of Time in UK",
      str_detect(variable, "disability_level") ~ "Level of Disability", 
      str_detect(variable, "unpaid_carers") ~ "Care Status",           
      str_detect(variable, "religion__") ~ "Religion",
      str_detect(variable, "economic_status") ~ "Economic Status",
      str_detect(variable, "segment") ~ "Segment",
      TRUE ~ "Other"
    )) %>%
    group_by(segment, category) %>%
    mutate(proportion = count / sum(count) * 100) %>%  # Ensure 100% sum per category
    select(segment, variable, category, proportion)
  
  return(data_long)
}

```

```{r}
# Apply function to both datasets
scotland_summary <- calculate_normalized_proportions_scotland(scotland) %>% mutate(region = "Scotland")
england_summary <- calculate_normalized_proportions_england(england_wales) %>% mutate(region = "England & Wales")

# Combine for comparison
comparison_summary <- bind_rows(scotland_summary, england_summary)

comparison_summary <- comparison_summary %>%
  mutate(proportion = round(proportion, 2))  # Round to 2 decimal places



```





calculate_normalized_proportions <- function(data) {
  # Step 1: Aggregate counts by segment
  data_long <- data %>%
    group_by(segment) %>%
    summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>%
    pivot_longer(cols = -segment, names_to = "variable", values_to = "count") %>%
    # Step 2: Assign each variable to a category
    mutate(category = case_when(
      str_detect(variable, "occupation|Manager|Process") ~ "Occupation",  
      str_detect(variable, "qualifications|Apprenticeship") ~ "Qualifications",
      str_detect(variable, "Owned|Shared|rented|Lives") ~ "Housing Tenure",
      str_detect(variable, "Household_") ~ "Household Deprivation Status",
      str_detect(variable, "Single_family_household|One_person_household|Multi_person_household|Other_household_types") ~ "Household Type",  
      str_detect(variable, "Male|Female") ~ "Sex",
      str_detect(variable, "national|identity|identities") ~ "National Identity",
      str_detect(variable, "Asian|Black|Mixed|White|Other_ethnic|Any_other_ethnic") ~ "Ethnicity",
      str_detect(variable, "Economically") ~ "Economic Status",
      str_detect(variable, "Europe|Africa|Middle_East|The_Americas|British_Overseas|Antarctica") ~ "Place of Birth",
      str_detect(variable, "Married|Divorced|Separated|Widowed|Never_married") ~ "Marital Status",
      str_detect(variable, "deprivation") ~ "Household Deprivation Status",
      str_detect(variable, "Welsh") ~ "Welsh Language",
      str_detect(variable, "Not_in") ~ "Employment Status",
      str_detect(variable, "health") ~ "Health Status", 
      str_detect(variable, "Christian|Buddhist|Hindu|Jewish|Muslim|Sikh|Other_religion|Not_answered|No_religion") ~ "Religion",
      str_detect(variable, "All_houshold_members|Ethnic_groups_differ|_same_ethnic_group") ~ "Household Ethnic Composition",
      str_detect(variable, "segment") ~ "Segment",
      str_detect(variable, "Aged") ~ "Age",
      TRUE ~ "Other"
    ))

```

```{r}
# Apply function to both datasets
scotland_summary <- calculate_normalized_proportions(scotland) %>% mutate(region = "Scotland")
england_summary <- calculate_normalized_proportions(england_wales) %>% mutate(region = "England & Wales")

# Combine for comparison
comparison_summary <- bind_rows(scotland_summary, england_summary)




```





```{r}

# Function to calculate proportions by segment
calculate_demographic_proportions <- function(data) {
  data %>%
    group_by(segment) %>%  # Group by segment
    summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>%  # Calculate mean proportions
    pivot_longer(cols = -segment, names_to = "variable", values_to = "proportion")  # Convert to long format
}

```

```{r}
# Apply function to both datasets
scotland_summary <- calculate_demographic_proportions(scotland) %>% mutate(region = "Scotland")
england_summary <- calculate_demographic_proportions(england_wales) %>% mutate(region = "England & Wales")

# Combine both for comparison
comparison_summary <- bind_rows(scotland_summary, england_summary)




```



```{r}
# Recode age variables in England & Wales to match Scotland's categories
england_wales <- england_wales %>%
  mutate(
    age__0_15 = Aged_4_years_and_under + Aged_5_to_9_years + Aged_10_to_14_years,
    age__16_24 = Aged_15_to_19_years + Aged_20_to_24_years,
    age__25_34 = Aged_25_to_29_years + Aged_30_to_34_years,
    age__35_49 = Aged_35_to_39_years + Aged_40_to_44_years + Aged_45_to_49_years,
    age__50_64 = Aged_50_to_54_years + Aged_55_to_59_years + Aged_60_to_64_years,
    age__65_and_over = Aged_65_to_69_years + Aged_70_to_74_years + Aged_75_to_79_years + 
                       Aged_80_to_84_years + Aged_85_years_and_over
  )

```

```{r}

## recode ethnicity## 

england_wales <- england_wales %>%
  mutate(
    ethnicity__white = White
    ethnicity__asian = Indian + Pakistani + Bangladeshi + Chinese + Other_Asian,
    ethnicity__black = Black_African + Black_Caribbean + Other_Black,
    ethnicity__mixed = Mixed_White_and_Black_Caribbean + Mixed_White_and_Black_African + Mixed_White_and_Asian + Other_Mixed,
    ethnicity__other = Other_ethnicity
  )

```











```{r}
## recode natioanl identity variables in England & Wales to match Scotland's categories

england_wales <- england_wales %>%
  mutate(
    national_identity__british = British,
    national_identity__english = English,
    national_identity__welsh = Welsh,
    national_identity__other = Other_national_identity
  )

```